---
title: 'MAT374: Chapter 2 homework'
author: PUT YOUR NAME HERE
date: \today
output: pdf_document
geometry: margin=0.85in
fontsize: 12pt
header-includes:
- \usepackage{setspace}\onehalfspacing
---

\tableofcontents
\listoffigures
\listoftables 

\newpage

```{r, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 5, echo = TRUE, eval = TRUE)
library(knitr)
library(tidyverse)


#library(tidyverse)
library(corrplot)
library(caret)

#data <- read.csv("used_cars_data.csv", stringsAsFactors = FALSE)

```


## Introduction

This project focuses on exploring the dynamics of the used car market by analyzing key factors that influence vehicle pricing. With the growing popularity of used vehicles, understanding how attributes such as brand, model, year, engine specifications, mileage, and kilometers driven affect the price is essential for both consumers and dealers. The dataset, `used_cars_data.csv`, sourced from Kaggle, provides comprehensive information on these characteristics, offering valuable insights into their contributions to car valuation. By leveraging this dataset, we aim to develop a robust predictive model that accurately estimates a used car's price based on its features. This research has practical applications, empowering buyers to make informed purchasing decisions and enabling sellers to price vehicles competitively in the marketplace.

Our second approach builds upon this analysis by employing a cross-validated linear regression model to enhance the accuracy of price predictions. This model will be compared to the initial model to evaluate improvements in predictive performance. The model is based on the general equation: `$Y = B_0 + B_1 x_1 + B_2 x_2 + B_n x_n$` where `$Y$` is the response variable(pricted price) and the `$B's$` represents the coefficients and the `$x's$` are the input variables. Key predictors include numeric variables such as mileage, engine capacity, and kilometers driven. Through rigorous pre-processing, including data cleaning and normalization, the model ensures the reliability and accuracy of predictions, providing deeper insights into the factors shaping the pricing of used cars.

Muhyadin:


This project aims to explore the factors influencing used car pricing in the growing market for pre-owned vehicles. Understanding how attributes like manufacturer, model, production year, engine type, and kilometers driven impact pricing is essential for making informed decisions. For consumers, this knowledge ensures fair deals, while for dealers, it aids in pricing strategies and inventory management. With the increasing demand for used vehicles, accurate pricing analysis benefits the entire market ecosystem.

This topic is especially relevant to us as we are from third world/developing countries where used cars dominate the automobile market due to their affordability. For many families and individuals, used cars represent a practical and economical choice, making it critical to understand the dynamics of their pricing. By analyzing these factors, we can help create a fairer, more transparent market for these essential vehicles.

The dataset, `used_cars_data.csv`, provides detailed information on these features, enabling a comprehensive study of their contributions to the sale price. To analyze this, we will develop a predictive linear regression model based on the equation $Y = B_0 + B_1 x_1 + B_2 x_2 +... + B_n x_p$, where $Y$ is the predicted price, $B$ represents coefficients, and $x$  represents independent variables. The process will involve data cleaning, normalization, and statistical modeling to ensure accuracy, readability and reliability.

By providing insights into the pricing dynamics of used cars, this project addresses a critical need in automobile markets in developing countries. It empowers buyers and sellers to navigate transactions with greater confidence, contributing to a more transparent and efficient market for pre-owned vehicles.

# Research Question
What are the key factors influencing the price of used cars and understanding how these factors can assist consumers in evaluating fair prices and help sellers optimize their pricing strategies


## Data Wrangling & Data Set Description 

# Data Set
The data set('used_cars_data.csv') from kaggle contains details of used cars, including price, year, kilometers driven, mileage, engine capacity, power, etc.


# Data wrangling
```{r}
# Data Cleaning
data$Mileage <- as.numeric(str_extract(data$Mileage, "\\d+\\.\\d+"))
data$Engine <- as.numeric(str_extract(data$Engine, "\\d+"))
data$Power <- as.numeric(str_extract(data$Power, "\\d+\\.\\d+"))

# Remove Rows with Missing Price
data <- data %>% filter(!is.na(Price))

# Select Relevant Variables
numerical_data <- data %>%
  select(Price, Year, Kilometers_Driven, Mileage, Engine, Power, Seats) %>%
  na.omit()
```
Data wrangling is a crucial step in the analysis process to prepare raw data for modeling and insights. In this case, the used car dataset contained several columns with numeric information stored as text (e.g., Mileage, Engine, and Power). These were converted to numeric formats to ensure they could be used in mathematical operations and modeling. Additionally, rows with missing values in the Price column were removed since price is the target variable, and missing values would interfere with predictions. Finally, non-essential columns were excluded, and rows with missing values in key variables were omitted to ensure the dataset was complete and consistent for analysis.

By performing these steps, we ensure that the data is clean, correctly formatted, and suitable for building reliable and interpretable predictive models. This process minimizes errors and ensures that our results are accurate and meaningful.

## Exploratory Data Analysis
Here is a visual representation of some of the variables.

```{r}

# Exploratory Data Analysis
# Correlation Matrix and Heatmap
cor_matrix <- cor(numerical_data)
corrplot(cor_matrix, method = "circle", type = "upper",
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix for Used Cars Data",
         mar = c(0, 0, 1, 0))

```
The correlation matrix plot, visualizing the relationships between numerical variables in the used car dataset. Each variable is paired with the others, and the size and color of the circles represent the strength and direction of their correlations. Positive correlations are shown in blue, while negative correlations are in red. The darker and larger the circle, the stronger the correlation. For instance, variables like Power and Price appear to have a strong positive correlation, indicating that as engine power increases, the price of the car tends to increase. Conversely, variables like Mileage show weaker or mixed relationships with other variables. This plot helps identify which features are most strongly related to the target variable (Price) for predictive modeling.

Muhyadin:

The correlation analysis of the dataset shows several key insights. Price and Power are positively correlated, which means that with increased power, the price of the vehicle goes up. Cars with higher horsepower generally command higher prices due to their enhanced performance and desirability. There is a moderate positive correlation between Year and Price, indicating that the price of newer car models is higher compared to older ones. In contrast, the Mileage is negatively correlated with Power and Engine, which indicates that high-mileage vehicles have low engine capacity and power. Lastly, Kilometers Driven is very poorly correlated with most of the variables, so this variable has very minimal direct impact on the pricing of used cars in this dataset. These findings provide valuable insights into the factors that influence used car prices.

```{r}
library(dplyr)
library(gganimate)
used_cars$Year <- as.numeric(used_cars$Year)

# Grouping data by Year and calculating the average Price
used_cars_time <- used_cars %>%
  group_by(Year) %>%
  summarise(avg_price = mean(Price, na.rm = TRUE))

# Plotting the average car price over time
ggplot(used_cars_time) +
 aes(x = Year, y = avg_price) +
 geom_point(color = "blue") +
 geom_line(color = "blue") +
 xlim(1980, 2024) +
 xlab("Year") +
 ylab("Average Car Price (USD)") +
 ggtitle("Figure: Average Car Price Over Time") +
 theme(plot.title = element_text(hjust = 0.5)) +
 transition_reveal(Year)
```

Explain with citation:


This graph shows that over the years, car prices have risen linearly, reflecting larger market trends and inflation. More specifically, there is a larger jump in average price after the year 2015. This may be due to the development of automotive technology as well as increased consumer desires for higher-end features. From the year 2000 to the year 2010, car prices remained somewhat similar, possibly due to the general global recession affecting the income of people. These trends suggest that economic factors, like inflation and market conditions, as well as technological innovations, make a big difference in how car prices change over time.

In India, car ownership was once seen as a luxury, and more importance was given to used cars. Used cars generally retain their value longer. During the early 2000s, the steady growth of India's economy and its moderate inflation levels supported consistent pricing, hence preventing extreme fluctuations in the automotive sector. However, car ownership was less common, and affordability remained a constraint, making demand and supply for used cars balanced and hence stable in terms of prices. This era thus illustrates how macroeconomic stability and shifting consumer perceptions together drive the used car market.



```{r}
# Histogram: Mileage Distribution
ggplot(numerical_data, aes(x = Mileage)) +
  geom_histogram(binwidth = 2, fill = "blue", alpha = 0.8) +
  labs(title = "Mileage Distribution", x = "Mileage (km/l)", y = "Count") +
  theme_minimal()
```
Explain with citation:

This graph shows that over the years, car prices have risen linearly, reflecting larger market trends and inflation. More specifically, there is a larger jump in average price after the year 2015. This may be due to the development of automotive technology as well as increased consumer desires for higher-end features. From the year 2000 to the year 2010, car prices remained somewhat similar, possibly due to the general global recession affecting the income of people. These trends suggest that economic factors, like inflation and market conditions, as well as technological innovations, make a big difference in how car prices change over time.

In India, car ownership was once seen as a luxury, and more importance was given to used cars. Used cars generally retain their value longer. During the early 2000s, the steady growth of India's economy and its moderate inflation levels supported consistent pricing, hence preventing extreme fluctuations in the automotive sector. However, car ownership was less common, and affordability remained a constraint, making demand and supply for used cars balanced and hence stable in terms of prices. This era thus illustrates how macroeconomic stability and shifting consumer perceptions together drive the used car market.


```{r}
# Bar Plot: Average Price by Seats
numerical_data %>%
  group_by(Seats) %>%
  summarise(Average_Price = mean(Price, na.rm = TRUE)) %>%
  ggplot(aes(x = as.factor(Seats), y = Average_Price)) +
  geom_bar(stat = "identity", fill = "blue", alpha = 0.8) +
  labs(title = "Average Price by Number of Seats", x = "Seats", y = "Average Price (Lakhs)") +
  theme_minimal()
```
Explain with citation:

The trends of car prices by seat capacity show a big difference. The average prices for two-seater cars are usually the highest, sometimes because they are luxury or sports models that attract premium buyers. On the other hand, five-seater cars, which are common family cars, are generally more reasonably priced due to their wide appeal and mass-market availability. While seven-seater cars show quite a remarkable rise in prices, reflecting their demand from larger families and the popularity of premium SUVs offering more space and comfort. However, vehicles with more than eight seats have a gradual decline in average prices; this means that these models are mostly utilitarian and rarely sell for luxury, mostly commercial or transport purposes. These trends indicate how functionality and market demand might affect pricing.


## Model and Results

# Model

```{r}
# Load necessary libraries
library(caret)

# Prepare the dataset
set.seed(123)  # For reproducibility
data <- numerical_data  # Use your cleaned numerical dataset
data <- na.omit(data)  # Ensure no missing values

# Define the formula
formula <- Price ~ Year + Kilometers_Driven + Mileage + Engine + Power + Seats

# Create a custom train control for LOOCV
control <- trainControl(method = "LOOCV")

# Train the model using LOOCV
model <- train(
  formula,
  data = data,
  method = "lm",  # Linear regression
  trControl = control
)

# Print the model summary
print(model)

# Make predictions for each LOOCV iteration
predictions <- predict(model, data)

# Compare predicted vs actual prices
results <- data.frame(
  Actual = data$Price,
  Predicted = predictions
)

# Print the first few rows of results
print(head(results))

# Calculate Mean Absolute Error (MAE)
mae <- mean(abs(results$Actual - results$Predicted))
cat("Mean Absolute Error (MAE):", mae, "\n")

ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", size = 3, alpha = 0.6) +  # Scatter points
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", size = 1) +  # Diagonal reference line
  labs(
    title = "Actual vs Predicted Prices for Used Cars",
    x = "Actual Price ",
    y = "Predicted Price "
  ) +
  theme_minimal() +  # Clean theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 14)
  )
```

Explain: The scatter plot illustrates the relationship between the actual and predicted prices of used cars based on the regression model. The blue points represent individual observations, while the red dashed line indicates the ideal case where predicted prices perfectly match actual prices (a 45-degree line). The clustering of points along the red line demonstrates that the model generally performs well, with many predictions close to their corresponding actual values. However, deviations from the line, particularly at higher price ranges, indicate areas where the model's predictions deviate from reality. For instance, some points above or below the line reflect overestimation or underestimation of car prices by the model. Overall, the model captures the general trend in pricing but shows increasing variance at higher price points, suggesting that additional factors or non-linear relationships might need to be considered to improve the accuracy of the predictions in these ranges.



# Results Summary
```{r}
# Print Model Summary

summary(model$finalModel)

```
Explain : The regression model predicts the dependent variable using six predictors: Year, Kilometers_Driven, Mileage, Engine, Power, and Seats. The residuals indicate a good fit, with the median close to zero (-0.492) and most residuals falling within a reasonable range (-2.520 to 1.789 for the interquartile range). Among the predictors, Year (β = 0.923, p < 2e-16), Kilometers_Driven (β = -1.102e-05, p = 0.000124), and Power (β = 0.164, p < 2e-16) have significant effects on the outcome variable, suggesting that newer models, lower mileage, and higher power are strongly associated with higher values of the outcome. However, Mileage, Engine, and Seats are not statistically significant (p > 0.05) and do not meaningfully contribute to the model. The model explains 69.4% of the variance in the dependent variable (R² = 0.6943), and the overall model is highly significant (F = 1395, p < 2.2e-16). This suggests that the model fits the data well, with Year, Power, and Kilometers_Driven being key predictors, but it may benefit from refinement by removing insignificant variables or exploring additional features.


# Model Performance
```{r}
# Load necessary libraries
library(dplyr)

# Read the datasets
used_cars <- read.csv("used_cars_data.csv", stringsAsFactors = FALSE)
cars93 <- read.csv("Cars93.csv", stringsAsFactors = FALSE)

# Inspect column names to identify possible keys for joining
cat("Columns in used_cars dataset:\n")
print(colnames(used_cars))

cat("\nColumns in Cars93 dataset:\n")
print(colnames(cars93))

# Check for overlap between potential join keys (e.g., 'Make', 'Model', 'Seats')
cat("\nUnique Seats in used_cars:\n")
print(unique(used_cars$Seats))

cat("\nUnique Passengers in Cars93:\n")
print(unique(cars93$Passengers))

# Rename column in Cars93 for joining (if necessary)
cars93 <- cars93 %>%
  rename(Seats = Passengers)

#cars93 <- cars93 %>% mutate(EngineSize = as.character(EngineSize * 1000))


cars93$'Min.Price' = NULL
cars93$'Max.Price' = NULL
cars93$'EngineSize' = NULL
cars93$'Make' = NULL

used_cars$'New_Price' = NULL
cars93$'Price.y' = NULL

# Perform a left join to retain all records from 'used_cars'
joined_data <- used_cars %>%
  left_join(cars93, by = "Seats")


# Ensure each observation appears only once by grouping by 'S.No' and selecting the first match
joined_data <- joined_data %>%
  group_by(S.No.) %>%
  slice(1) %>%
  ungroup()

# Inspect the joined dataset
cat("\nPreview of joined dataset:\n")
print(head(joined_data))

# Check dimensions of the joined dataset
cat("\nDimensions of the joined dataset:\n")
print(dim(joined_data))

# Perform exploratory analysis on the joined dataset
# Example: Summarize average price by Seats
summary_by_seats <- joined_data %>% 
  group_by(Seats)  
  #summarize(
    #Avg_Used_Car_Price = mean(Price.x, na.rm = TRUE)
  #)

# Output the summary
cat("\nSummary of average prices by Seats:\n")
print(summary_by_seats)


# Save the joined dataset to a CSV file
#write.csv(joined_data, "joined_cars_data.csv", row.names = FALSE)

# Save the summary to a CSV file
#write.csv(summary_by_seats, "summary_by_seats.csv", row.names = FALSE)

```
Explain with citations:


```{r}
library(dplyr)
library(stringr)

# Data Cleaning
# Extract numerical values from Mileage, Engine, and Power columns
joined_data <- joined_data %>%
  mutate(
    Mileage = as.numeric(str_extract(Mileage, "\\d+\\.\\d+")),
    Engine = as.numeric(str_extract(Engine, "\\d+")),
    Power = as.numeric(str_extract(Power, "\\d+\\.\\d+"))
    
  )

joined_data <- joined_data %>%
  rename(Price = Price.x)

# Remove rows with missing Price
joined_data <- joined_data %>%
  filter(!is.na(Price))

# Select relevant numerical variables and remove rows with missing values
joined_data <- joined_data %>%
  select(Price, Year, Kilometers_Driven, Mileage, Engine, Power, Seats) %>%
  na.omit()

# Load necessary libraries
#library(dplyr)
library(caret) # For model training and evaluation

# Inspect the data to ensure it is clean
cat("Preview of the dataset for modeling:\n")
print(head(joined_data))

# Check for missing values and handle them
cat("\nSummary of missing values:\n")
print(colSums(is.na(joined_data)))

# Impute or remove rows with missing Price values
joined_data <- joined_data %>% filter(!is.na(Price))

# Select relevant features for prediction
# Drop unnecessary columns and keep relevant ones for modeling
joined_data <- joined_data %>% 
  select(Price,  Year, Kilometers_Driven, Mileage, Engine, Power,  Seats,) %>% 
  na.omit()


# Rename Price column for consistency
colnames(joined_data)[1] <- "Price"

# Split the data into training and testing sets
set.seed(123)  # For reproducibility
train_index <- createDataPartition(joined_data$Price, p = 0.8, list = FALSE)
train_data <- joined_data[train_index, ]
test_data <- joined_data[-train_index, ]

# Train a linear regression model
cat("\nTraining the linear regression model...\n")
model <- train(
  Price ~ ., 
  data = train_data, 
  method = "lm"
)

# Display the model summary
cat("\nModel Summary:\n")
print(summary(model$finalModel))

# Make predictions on the test data
cat("\nMaking predictions on the test dataset...\n")
predictions <- predict(model, newdata = test_data)


# Evaluate the model performance
cat("\nModel Performance Metrics:\n")
rmse <- RMSE(predictions, test_data$Price)
r_squared <- R2(predictions, test_data$Price)

cat(paste0("Root Mean Squared Error (RMSE): ", round(rmse, 2), "\n"))
cat(paste0("R-squared: ", round(r_squared, 2), "\n"))

# Plot actual vs predicted values
plot(test_data$Price, predictions, 
     xlab = "Actual Price", 
     ylab = "Predicted Price", 
     main = "Actual vs Predicted Prices")
abline(0, 1, col = "red")  # Add a 45-degree line for reference
```
Explain with citation:


```{r}
# Clean the 'Engine' column to remove 'CC' and convert it to numeric
used_cars_data$Engine <- as.numeric(gsub(" CC", "", used_cars_data$Engine))

# Pivot to calculate average engine size by Fuel Type
fuel_type_avg_engine <- used_cars_data %>%
  group_by(Fuel_Type) %>%
  summarise(Average_Engine_Size = mean(Engine, na.rm = TRUE)) %>%
  arrange(desc(Average_Engine_Size))

# View the result
print(fuel_type_avg_engine)
```
```{r}
ggplot(fuel_type_avg_engine, aes(x = reorder(Fuel_Type, -Average_Engine_Size), y = Average_Engine_Size)) +
  geom_bar(stat = "identity", fill = "coral") +
  geom_text(aes(label = round(Average_Engine_Size, 1)), 
            vjust = -0.5, # Adjust vertical position of the labels
            color = "black", 
            size = 4) + # Adjust text size
  labs(title = "Average Engine Size by Fuel Type",
       x = "Fuel Type",
       y = "Average Engine Size (CC)") +
  theme_minimal()

```

Explain with citation:
```{r}
# Load necessary libraries
library(dplyr)
library(stringr)
library(caret) # For model training and evaluation

# Data Cleaning
# Extract numerical values from Mileage, Engine, and Power columns
joined_data <- joined_data %>%
  mutate(
    Mileage = as.numeric(str_extract(Mileage, "\\d+\\.\\d+")),
    Engine = as.numeric(str_extract(Engine, "\\d+")),
    Power = as.numeric(str_extract(Power, "\\d+\\.\\d+"))
  )

joined_data <- joined_data %>%
  rename(Price = Price.x)

# Remove rows with missing Price
joined_data <- joined_data %>%
  filter(!is.na(Price))

# Select relevant numerical variables and remove rows with missing values
joined_data <- joined_data %>%
  select(Price, Year, Kilometers_Driven, Mileage, Engine, Power, Seats) %>%
  na.omit()

# Inspect the data to ensure it is clean
cat("Preview of the dataset for modeling:\n")
print(head(joined_data))

# Check for missing values and handle them
cat("\nSummary of missing values:\n")
print(colSums(is.na(joined_data)))

# Split the data into training and testing sets
set.seed(123)  # For reproducibility
train_index <- createDataPartition(joined_data$Price, p = 0.8, list = FALSE)
train_data <- joined_data[train_index, ]
test_data <- joined_data[-train_index, ]

# Train a linear regression model
cat("\nTraining the linear regression model...\n")
model <- train(
  Price ~ ., 
  data = train_data, 
  method = "lm"
)

# Display the model summary
cat("\nModel Summary:\n")
print(summary(model$finalModel))

# Make predictions on the test data
cat("\nMaking predictions on the test dataset...\n")
predictions <- predict(model, newdata = test_data)

# Evaluate the model performance
cat("\nModel Performance Metrics:\n")
rmse <- RMSE(predictions, test_data$Price)
r_squared <- R2(predictions, test_data$Price)

cat(paste0("Root Mean Squared Error (RMSE): ", round(rmse, 2), "\n"))
cat(paste0("R-squared: ", round(r_squared, 2), "\n"))

# Plot actual vs predicted values
plot(test_data$Price, predictions, 
     xlab = "Actual Price", 
     ylab = "Predicted Price", 
     main = "Actual vs Predicted Prices")
abline(0, 1, col = "red")  # Add a 45-degree line for reference

```
Explain with citations:

## Ethical Considerations



## Conclusion









`
