---
title: "Project"
output:
  html_document:
    df_print: paged
date: "2024-12-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(corrplot)
library(caret)

```


## Introduction

This project focuses on exploring the dynamics of the used car market by analyzing key factors that influence vehicle pricing. With the growing popularity of used vehicles, understanding how attributes such as brand, model, year, engine specifications, mileage, and kilometers driven affect the price is essential for both consumers and dealers. The dataset, `used_cars_data.csv`, sourced from Kaggle, provides comprehensive information on these characteristics, offering valuable insights into their contributions to car valuation. By leveraging this dataset, we aim to develop a robust predictive model that accurately estimates a used car's price based on its features. This research has practical applications, empowering buyers to make informed purchasing decisions and enabling sellers to price vehicles competitively in the marketplace.

Our second approach builds upon this analysis by employing a cross-validated linear regression model to enhance the accuracy of price predictions. This model will be compared to the initial model to evaluate improvements in predictive performance. The model is based on the general equation: `$Y = B_0 + B_1 x_1 + B_2 x_2 + B_n x_n$` where `$Y$` is the response variable(pricted price) and the `$B's$` represents the coefficients and the `$x's$` are the input variables. Key predictors include numeric variables such as mileage, engine capacity, and kilometers driven. Through rigorous pre-processing, including data cleaning and normalization, the model ensures the reliability and accuracy of predictions, providing deeper insights into the factors shaping the pricing of used cars.


# Research Question
What are the key factors influencing the price of used cars and understanding how these factors can assist consumers in evaluating fair prices and help sellers optimize their pricing strategies


## Data Wrangling & Data Set Description 

# Data Set
The data set('used_cars_data.csv') from kaggle contains details of used cars, including price, year, kilometers driven, mileage, engine capacity, power, etc.


# Data wrangling
```{r}
# Load necessary libraries
library(dplyr)

# Read the datasets
used_cars <- read.csv("used_cars_data.csv", stringsAsFactors = FALSE)
cars93 <- read.csv("Cars93.csv", stringsAsFactors = FALSE)

# Inspect column names to identify possible keys for joining
cat("Columns in used_cars dataset:\n")
print(colnames(used_cars))

cat("\nColumns in Cars93 dataset:\n")
print(colnames(cars93))

# Check for overlap between potential join keys (e.g., 'Make', 'Model', 'Seats')
cat("\nUnique Seats in used_cars:\n")
print(unique(used_cars$Seats))

cat("\nUnique Passengers in Cars93:\n")
print(unique(cars93$Passengers))

# Rename column in Cars93 for joining (if necessary)
cars93 <- cars93 %>%
  rename(Seats = Passengers)

cars93 <- cars93 %>% mutate(EngineSize = as.character(EngineSize * 1000))


cars93$'Min.Price' = NULL
cars93$'Max.Price' = NULL
cars93$'EngineSize' = NULL
cars93$'Make' = NULL

used_cars$'New_Price' = NULL
cars93$'Price.y' = NULL

# Perform a left join to retain all records from 'used_cars'
joined_data <- used_cars %>%
  left_join(cars93, by = "Seats")


# Ensure each observation appears only once by grouping by 'S.No' and selecting the first match
joined_data <- joined_data %>%
  group_by(S.No.) %>%
  slice(1) %>%
  ungroup()

# Inspect the joined dataset
cat("\nPreview of joined dataset:\n")
print(head(joined_data))

# Check dimensions of the joined dataset
cat("\nDimensions of the joined dataset:\n")
print(dim(joined_data))

# Perform exploratory analysis on the joined dataset
# Example: Summarize average price by Seats
summary_by_seats <- joined_data %>% 
  group_by(Seats) %>% 
  summarize(
    Avg_Used_Car_Price = mean(Price.x, na.rm = TRUE)
  )

# Output the summary
cat("\nSummary of average prices by Seats:\n")
print(summary_by_seats)


# Save the joined dataset to a CSV file
#write.csv(joined_data, "joined_cars_data.csv", row.names = FALSE)

# Save the summary to a CSV file
#write.csv(summary_by_seats, "summary_by_seats.csv", row.names = FALSE)

```

```{r}
library(dplyr)
library(tidyr)

# Create a pivot table in long format
pivot_long <- joined_data %>%
  pivot_longer(
    cols = starts_with("Price"), # Choose columns to pivot (e.g., "Price.x" and "Price.y")
    names_to = "Price_Type",     # New column to store names of pivoted columns
    values_to = "Price"          # New column to store the values
  )

# Preview the long format data
#cat("\nPivoted data in long format:\n")
print(head(pivot_long))

# Save the long format data to a CSV file
#write.csv(pivot_long, "pivot_long.csv", row.names = FALSE)

```

```{r}
# Create a pivot table in wide format
pivot_wide <- joined_data %>%
  pivot_wider(
    names_from = Seats,       # Column to create new column names
    values_from = Price.x     # Column to fill in as values
  )

# Preview the wide format data
cat("\nPivoted data in wide format:\n")
print(head(pivot_wide))

# Save the wide format data to a CSV file
write.csv(pivot_wide, "pivot_wide.csv", row.names = FALSE)

```
```{r}
library(dplyr)
library(stringr)

# Data Cleaning
# Extract numerical values from Mileage, Engine, and Power columns
joined_data <- joined_data %>%
  mutate(
    Mileage = as.numeric(str_extract(Mileage, "\\d+\\.\\d+")),
    Engine = as.numeric(str_extract(Engine, "\\d+")),
    Power = as.numeric(str_extract(Power, "\\d+\\.\\d+"))
  )

joined_data <- joined_data %>%
  rename(Price = Price.x)

# Remove rows with missing Price
data <- joined_data %>%
  filter(!is.na(Price))

# Select relevant numerical variables and remove rows with missing values
numerical_data <- data %>%
  select(Price, Year, Kilometers_Driven, Mileage, Engine, Power, Seats) %>%
  na.omit()

# Preview the cleaned data
cat("\nPreview of cleaned numerical data:\n")
print(head(numerical_data))

# Check dimensions of the cleaned dataset
cat("\nDimensions of the cleaned numerical data:\n")
print(dim(numerical_data))

```
```{r}

# Exploratory Data Analysis
# Correlation Matrix and Heatmap
cor_matrix <- cor(numerical_data)
corrplot(cor_matrix, method = "circle", type = "upper",
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix for Used Cars Data",
         mar = c(0, 0, 1, 0))

```
The correlation matrix plot, visualizing the relationships between numerical variables in the used car dataset. Each variable is paired with the others, and the size and color of the circles represent the strength and direction of their correlations. Positive correlations are shown in blue, while negative correlations are in red. The darker and larger the circle, the stronger the correlation. For instance, variables like Power and Price appear to have a strong positive correlation, indicating that as engine power increases, the price of the car tends to increase. Conversely, variables like Mileage show weaker or mixed relationships with other variables. This plot helps identify which features are most strongly related to the target variable (Price) for predictive modeling.


```{r}
library(dplyr)
library(gganimate)
used_cars$Year <- as.numeric(used_cars$Year)

# Grouping data by Year and calculating the average Price
used_cars_time <- joined_data %>%
  group_by(Year) %>%
  summarise(avg_price = mean(Price, na.rm = TRUE))

# Plotting the average car price over time
ggplot(used_cars_time) +
 aes(x = Year, y = avg_price) +
 geom_point(color = "blue") +
 geom_line(color = "blue") +
 xlim(1980, 2024) +
 xlab("Year") +
 ylab("Average Car Price (USD)") +
 ggtitle("Figure: Average Car Price Over Time") +
 theme(plot.title = element_text(hjust = 0.5)) +
 transition_reveal(Year)
```


```{r}
# Histogram: Mileage Distribution
ggplot(numerical_data, aes(x = Mileage)) +
  geom_histogram(binwidth = 2, fill = "blue", alpha = 0.8) +
  labs(title = "Mileage Distribution", x = "Mileage (km/l)", y = "Count") +
  theme_minimal()
```


```{r}
# Bar Plot: Average Price by Seats
numerical_data %>%
  group_by(Seats) %>%
  summarise(Average_Price = mean(Price, na.rm = TRUE)) %>%
  ggplot(aes(x = as.factor(Seats), y = Average_Price)) +
  geom_bar(stat = "identity", fill = "blue", alpha = 0.8) +
  labs(title = "Average Price by Number of Seats", x = "Seats", y = "Average Price ") +
  theme_minimal()
```
```{r}
# Clean the 'Engine' column to remove 'CC' and convert it to numeric
joined_data$Engine <- as.numeric(gsub(" CC", "", joined_data$Engine))

# Pivot to calculate average engine size by Fuel Type
fuel_type_avg_engine <- joined_data %>%
  group_by(Fuel_Type) %>%
  summarise(Average_Engine_Size = mean(Engine, na.rm = TRUE)) %>%
  arrange(desc(Average_Engine_Size))

# View the result
print(fuel_type_avg_engine)
```

```{r}
ggplot(fuel_type_avg_engine, aes(x = reorder(Fuel_Type, -Average_Engine_Size), y = Average_Engine_Size)) +
  geom_bar(stat = "identity", fill = "coral") +
  geom_text(aes(label = round(Average_Engine_Size, 1)), 
            vjust = -0.5, # Adjust vertical position of the labels
            color = "black", 
            size = 4) + # Adjust text size
  labs(title = "Average Engine Size by Fuel Type",
       x = "Fuel Type",
       y = "Average Engine Size (CC)") +
  theme_minimal()
```

```{r}
# Load necessary libraries
library(caret)
library(ggplot2)

# Prepare the dataset
set.seed(123)  # For reproducibility
data <- joined_data  # Use your cleaned numerical dataset
data <- na.omit(data)  # Ensure no missing values

# Define the formula
formula <- Price ~ Year + Kilometers_Driven + Mileage + Engine + Power + Seats

# Create a custom train control for LOOCV
control <- trainControl(method = "LOOCV")

# Train the model using LOOCV
model <- train(
  formula,
  data = data,
  method = "lm",  # Linear regression
  trControl = control
)

# Print the model summary
print(model)

# Make predictions for each LOOCV iteration
predictions <- predict(model, data)

# Compare predicted vs actual prices
results <- data.frame(
  Actual = data$Price,
  Predicted = predictions
)

# Print the first few rows of results
print(head(results))

# Calculate Mean Absolute Error (MAE)
mae <- mean(abs(results$Actual - results$Predicted))
cat("Mean Absolute Error (MAE):", mae, "\n")

# Plot Actual vs Predicted Prices
ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", size = 3, alpha = 0.6) +  # Scatter points
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", size = 1) +  # Diagonal reference line
  labs(
    title = "Actual vs Predicted Prices for Used Cars",
    x = "Actual Price",
    y = "Predicted Price"
  ) +
  theme_minimal() +  # Clean theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 14)
)
```


